Review of Edge Detectors

### 2014-11-11 old_versions/edge-detection-review-v9.tex
second round of third revision, improving references

%-------------------------------------------------------------------------------
% A Review of Classic Edge Detectors - v10
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------

@@

In 1980 Marr and Hildreth \cite{AIM-518} argued that intensity changes are not independent 
of image scale, so edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. This algorithm is presented in Section \ref{sec:second}. 
Haralick's algorithm \cite{bb20239} is an alternative approach based on the second derivative 
which is also reviewed in section \ref{sec:second}. This algorithm has the particularity of 
proposing a model to approximate locally the image around a point. Then, using this model, 
an approximation to the second derivative of the image can be calculated analytically and 
finding edges is achieved by imposing a condition over the model parameters. \\

@@

Finally, the edges image is obtained by thresholding the gradient magnitude image, i.e.\ for pixel at position $i$ in the image compute
\begin{equation*}
	\mbox{edges}[i] = \begin{cases} 0,& \mbox{if}\ M[i]\geq\mbox{th} \\
									1,& \mbox{if}\ M[i]<\mbox{th}
					  \end{cases}
\end{equation*}
where $\mbox{th}$ is the threshold. A black and white image is obtained, in which edge points are indicated in white (see Figure \ref{fig:thresholding}). 
It is noted that different thresholds lead to different results. Thicker or more edges for smaller thresholds, fewer edges 
for larger thresholds. This threshold, and the others mentioned below, are defined as a percentage of the maximum value 
of the gradient image. The advantage of this is to adapt the threshold to the dynamic range of the image. 
Therefore, the parameter $threshold$ in the algorithms take values ​​between $0$ and $1$.

@@

where $\rho_0$ is a fixed threshold, passed as an argument to the algorithm. Ideally, the polynomial function 
should be zero at $\rho=0$, but this condition is relaxed, allowing the the polynomial becomes zero in a neighborhood 
of $\rho=0$ ($\rho\leq\rho_0$). This parameter must be greater than zero and and less than $\sqrt{2}$. Its optimal 
value is between $0.4$ and $0.6$.\\

@@

\begin{equation}
	\label{eq:log_function}
	LoG(x,y)=\frac{1}{\sqrt{2\pi\sigma^2}}\frac{x^2+y^2-2\sigma^2}{\sigma^4}e^{-(x^2+y^2)/2\sigma^2}.
\end{equation}\\

@@

	\subfigure[Marr-Hildreth (Gaussian). $\sigma=1.5$, $n=13$, $th_{ZC}=0.1$.]{\label{fig:result1-d}\includegraphics[width=0.2\textwidth]{results/square127_marr-hildreth.png}}
	\quad
	\subfigure[Marr-Hildreth (LoG). $\sigma=1.5$, $n=17$, $th_{ZC}=0.1$.]{\label{fig:result1-e}\includegraphics[width=0.2\textwidth]{results/square127_marr-hildreth-log.png}}
	\quad
	\subfigure[Haralick. $\rho=0.4$.]{\label{fig:result1-f}\includegraphics[width=0.2\textwidth]{results/square127_haralick.png}}
	\caption{Results of the algorithms using the Figure \ref{fig:original1} as input image. The first three (\subref{fig:result1-a}, \subref{fig:result1-b} and \subref{fig:result1-c}) correspond to the first derivative methods (Roberts, Prewitt and Sobel), then the following two (\subref{fig:result1-d} and \subref{fig:result1-e}) are from the Marr-Hildreth algorithm using Gaussian and LoG kernels, and the last one (\subref{fig:result1-f}) corresponds to the Haralick algorithm. Note the rounded corners on the Marr-Hildreth results, due to Gaussian blur.}
	\label{fig:result1}
\end{figure}

\begin{table}[h!]
	\begin{center}
	\begin{tabular}{| l | r |}
		\hline \rule{0pt}{3ex}
		\cellcolor[gray]{0.8} \textbf{Algorithm}	& \cellcolor[gray]{0.8} \textbf{Execution time (s)}	\\ \hline \rule{0pt}{3ex}
		Roberts, Prewitt and Sobel					& $0.020 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (Gaussian)					& $0.030 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (LoG)							& $0.050 \ s$										\\ \hline \rule{0pt}{3ex}
		Haralick									& $0.040 \ s$										\\
		\hline
	\end{tabular}
	\end{center}
	\caption{Execution time of the algorithms (including I/O) using the Figure \ref{fig:original1} as input image (128$\times$128 pixels, 1 channel).}
	\label{exectime1}
\end{table}
\vspace{0.5cm}

@@

Some observations: 
\begin{itemize}
	\item The first derivative algorithms, although they work properly and quickly, have some problems like discontinuity of the edges. They are also very affected by noise in images, appearing unwanted isolated edges (More sophisticated methods help to avoid and improve this, e.g. Canny edge detector \cite{Canny1986}). \\
	\item The Marr-Hildreth algorithm (in its two versions) achieved interesting results in the edge detail. This algorithm provides a first attempt to obtain regular edges, because of the filtering with a Gaussian kernel. This behavior can be seen in Lena's hair (Figure \ref{fig:example1}), or inside the oranges (example \ref{fig:example2}). \\
	\item As mentioned earlier, Haralick's algorithm is the first to assume greater regularity of the image in the neighborhood of a pixel (third order). This causes thicker edges, as shown in both examples. But, note that some edges are only detected with this algorithm, e.g. the lower edges of the oranges, which are in the shadow, in Figure \ref{fig:example2}. These edges are quite smooth, so that the Haralick model is well suited to them, and they do not represent an abrupt change in the intensity to be detected by other methods. \\
\end{itemize}

@@

Haralick's algorithm, although it shares the Marr-Hildreth idea of finding zero crossings of the second derivative, has some quite different ideas. It is the only one of these algorithms which works with an approximation of the intensity of the image in the neighborhood of a pixel, using a bicubic polinomial function. This supposes a fairly large regularity in the image, which is not always true, and sometimes leads to detect edges where none of them exist, and get some thicker edges too. \\

All algorithms have their advantages and disadvantages. Choosing one or the other may depend on requirements of the application. Furthermore, none of the implemented methods ensure edge connectivity, as the Canny \cite{Canny1986} algorithm does.\\

@@

As an addendum to this work a detailed implementation of the described algorithms is presented, 
available in C language, along with an online demo where users can run the algorithms 
with arbitrary images uploaded by themselves. Also some code documentation tools are supplied, 
which were used to generate the detailed code description presented in section \ref{sec:appendix2}. We think this tool could be useful for future IPOL articles.

>>>>

%-------------------------------------------------------------------------------
% A Review of Classic Edge Detectors - v9
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------

@@

In 1980 Marr and Hildreth \cite{AIM-518} argued that intensity changes are not independent 
of image scale, so edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. This algorithm is presented in Section \ref{sec:second}. 
Haralick's algorithm \cite{bb20239} is an alternative approach based on the second derivative 
which is also reviewed in section \ref{sec:second}. This algorithm has the particularity of 
proposing a model to locally approximate the image around a point. Then, using this model, 
an approximation to the second derivative of the image can be calculated analytically and 
finding edges is done by imposing a condition over the model parameters. \\

@@

Finally, the edges image is obtained by thresholding the gradient magnitude image, i.e.\ for each pixel in the image compute
\begin{equation*}
	\mbox{edges}[i] = \begin{cases} 0,& \mbox{if}\ M[i]\geq\mbox{threshold} \\
									1,& \mbox{if}\ M[i]<\mbox{threshold}.
					  \end{cases}
\end{equation*}
A black and white image is obtained, in which edge points are indicated in white (see Figure \ref{fig:thresholding}). 
It is noted that different thresholds lead to different results. Thicker edges for smaller thresholds, lack of edges 
for larger thresholds. This threshold, and the others mentioned below, are defined as a percentage of the maximum value 
of the image to be processed. The advantage of this is to adapt the threshold to the dynamic range of the image. 
Therefore, the parameter $threshold$ in the algorithms take values ​​between $0$ and $1$.

@@

where $\rho_0$ is a fixed threshold, passed as an argument to the algorithm. This parameter indicates how far 
from the center of the pixel can be a zero of the second derivative. It must be greater than zero and and 
less than $\sqrt{2}$. Its optimal value is between $0.4$ and $0.6$.\\

@@

\begin{equation}
	\label{eq:log_function}
	LoG(x,y)=\frac{x^2+y^2-2\sigma^2}{\sigma^4}e^{-(x^2+y^2)/2\sigma^2}.
\end{equation}\\

@@

	\subfigure[Marr-Hildreth (Gaussian). $\sigma=1.5$, $n=13$, $th_{ZC}=0.1$.]{\label{fig:result1-d}\includegraphics[width=0.2\textwidth]{results/square127_marr-hildreth.png}}
	\quad
	\subfigure[Marr-Hildreth (LoG). $\sigma=1.5$, $n=17$, $th_{ZC}=0.1$.]{\label{fig:result1-e}\includegraphics[width=0.2\textwidth]{results/square127_marr-hildreth-log.png}}
	\quad
	\subfigure[Haralick. $\rho=0.4$.]{\label{fig:result1-f}\includegraphics[width=0.2\textwidth]{results/square127_haralick.png}}
	\caption{Results of the algorithms using the Figure \ref{fig:original1} as input image. The first three (\ref{fig:result1-a}, \ref{fig:result1-b} and \ref{fig:result1-c}) correspond to the first derivative methods (Roberts, Prewitt and Sobel), then the following two (\ref{fig:result1-d} and \ref{fig:result1-e}) are from the Marr-Hildreth algorithm using Gaussian and LoG kernels, and the last one (\ref{fig:result1-f}) corresponds to the Haralick algorithm. Note the rounded corners on the Marr-Hildreth results, due to Gaussian blur.}
	\label{fig:result1}
\end{figure}

\begin{table}[t!]
	\begin{center}
	\begin{tabular}{| l | r |}
		\hline \rule{0pt}{3ex}
		\cellcolor[gray]{0.8} \textbf{Algorithm}	& \cellcolor[gray]{0.8} \textbf{Execution time (s)}	\\ \hline \rule{0pt}{3ex}
		Roberts, Prewitt and Sobel					& $0.020 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (Gaussian)					& $0.030 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (LoG)							& $0.050 \ s$										\\ \hline \rule{0pt}{3ex}
		Haralick									& $0.040 \ s$										\\
		\hline
	\end{tabular}
	\end{center}
	\caption{Execution time of the algorithms (including i/o) using the Figure \ref{fig:original1} as input image (128$\times$128 pixels, 1 channel).}
	\label{exectime1}
\end{table}
\vspace{0.5cm}

@@

Some observations: 
\begin{itemize}
	\item The first derivative algorithms, although they work properly and quickly, have some problems like discontinuity of the edges. They are also very affected by noise in images, appearing unwanted isolated edges (More sophisticated methods help to avoid and improve this, e.g. Canny edge detector \cite{Canny1986}). \\
	\item The Marr-Hildreth algorithm (in its two versions) achieved interesting results in the edge detail. This algorithm provides a first step to regularity in the image, because of the filtering with a Gaussian kernel. This behavior can be seen in Lena's hair (Figure \ref{fig:example1}), or inside the oranges (example \ref{fig:example2}). \\
	\item As mentioned earlier, Haralick's algorithm is the first to assume greater regularity of the image in the neighborhood of a pixel (third order). This causes thicker edges, as shown in both examples. But, note that some edges are only detected with this algorithm, e.g. the lower edges of the oranges, which are in the shadow, in Figure \ref{fig:example2}. These edges are quite smooth, so that the Haralick model is well suited to them, and they do not represent an abrupt change in the intensity to be detected by other methods. \\
\end{itemize}

@@

Haralick's algorithm, although it shares the Marr-Hildreth idea of finding zero crossings of the second derivative, has some quite different ideas. It is the only one of these algorithms which works with an approximation of the intensity of the image in the neighborhood of a pixel, using a bicubic polinimial function. This supposes a fairly large regularity in the image, which is not always true, and sometimes leads to detect edges where none of them exist, and get some thicker edges too. \\

All algorithms have their advantages and disadvantages. Choosing one or the other may depend on requirements of the application. \\

@@

As an addendum to this work a detailed implementation of the described algorithms is presented, 
available in C language, along with an online demo where users can run the algorithms 
with arbitrary images uploaded by themselves. Also some code documentation tools are supplied, 
which were used to generate the detailed code description presented in section \ref{sec:appendix2} (Appendix 2). We think this tool could be useful for future IPOL articles.

### 2014-11-11 old_versions/edge-detection-review.tex
second round of third revision, improving references

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\begin{document}

@@

%-------------------------------------------------------------------------------
\begin{ipolAbstract}
In this paper we study different alternatives for edge detection in digital images. The main idea 
behind the edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts. The second derivative is also used, for example in algorithms 
like Marr-Hildreth and Haralick.\\
We compare the results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time), considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).\\

@@

Edge detection is the first approach for segmenting images
based on abrupt changes in intensity. Detecting those changes in intensity 
for the purpose of finding edges in images can be accomplished using first 
or second order derivatives.

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image. In section ~\ref{sec:first} we study the 
performance of these algorithms, as an introduction to a more sophisticated 
analysis into the edge finding process.\\

@@

Let us analyze the performance of each of the methods described above using some test images. The first image 
is a simple example, a white square on black background. In figure \ref{fig:result1} we can see the results 
of applying each of the algorithms to this image.

>>>>

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\numberwithin{figure}{section}
\begin{document}

@@

%-------------------------------------------------------------------------------
\begin{ipolAbstract}
In this paper we study some of the classic alternatives for edge detection in digital images. The main idea 
behind the edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts. The second derivative is also used, for example in algorithms 
like Marr-Hildreth and Haralick.\\
We compare the results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time), considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).\\

@@

Edge detection is one of the first approaches published in the literature for 
segmenting images. The basic idea is to detect abrupt changes in intensity. 
Detecting those changes in intensity for the purpose of finding edges in images 
can be accomplished using first or second order derivatives.

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image. In section ~\ref{sec:first} we study the 
performance of such algorithms, as an introduction to a more sophisticated 
analysis into the edge finding process.\\

@@

Let us analyze the performance of each of the methods described above using a test image shown in figure \ref{fig:original}.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.7\textwidth]{results/molino.eps}
	\caption{Test image: windmill.}
	\label{fig:original}
\end{figure}

First we analyze the results of the first derivative edge detection algorithms.

\begin{figure}[h!]
	\centering
	\subfloat[Roberts.]{\label{fig:result1-a}\includegraphics[width=0.5\textwidth]{results/molino.roberts.eps}}

	\subfloat[Prewitt.]{\label{fig:result1-b}\includegraphics[width=0.5\textwidth]{results/molino.prewitt.eps}}

	\subfloat[Sobel.]{\label{fig:result1-c}\includegraphics[width=0.5\textwidth]{results/molino.sobel.eps}}
	\caption{Results of the algorithms using the figure \ref{fig:original} as input image}
	\label{fig:result1}
\end{figure}

%\begin{figure}[b!]
%	\centering
%	\subfloat[Roberts.]{\label{fig:result1-a}\includegraphics[width=0.3\textwidth]{results/square/square.roberts.eps}}
%	\quad
%	\subfloat[Prewitt.]{\label{fig:result1-b}\includegraphics[width=0.3\textwidth]{results/square/square.prewitt.eps}}
%	\quad
%	\subfloat[Sobel.]{\label{fig:result1-c}\includegraphics[width=0.3\textwidth]{results/square/square.sobel.eps}}

%	\subfloat[Marr-Hildreth (Gaussian).]{\label{fig:result1-a}\includegraphics[width=0.3\textwidth]{results/square/square.marr-hildreth.eps}}
%	\quad
%	\subfloat[Marr-Hildreth (LoG).]{\label{fig:result1-b}\includegraphics[width=0.3\textwidth]{results/square/square.marr-hildreth-log.eps}}
%	\quad
%	\subfloat[Haralick.]{\label{fig:result1-c}\includegraphics[width=0.3\textwidth]{results/square/square.haralick.eps}}
%	\caption{Results of the algorithms using the figure \ref{fig:original1} as input image}
%	\label{fig:result1}
%\end{figure}
